---
title: "doomsayer_diagnostics"
output:
  html_document:
      toc: true
      toc_float: true
      keep_md: yes
      theme: united
  pdf_document:
date: '`r format(Sys.Date())`'
params:
  dev: no
  yaml_cfg: x
---

<style>
body {
    position: absolute;
    left: 0px;}
</style>

```{r child="R/setup.Rmd"}
```

### Error profile heatmap

Left panel shows root mean squared error (RMSE) of each outlier sample's mutation spectrum. Right panel shows the fold-difference in contribution per subtype for each outlier. Fold-differences are calculated relative to the mean contribution of that subtype across all samples.

```{r child="R/heatmap.Rmd"}
```

### Combined mutation spectrum

These plots shows the total number of observations in each subtype, combined across all samples

#### All samples

```{r child="R/spectra.Rmd"}
```

### Signature loadings (H matrix)

Describes how each mutation subtype is loaded into the r signatures

```{r child="R/sigloads.Rmd"}
```


```{r child="R/sigloads_bar.Rmd"}
```

### Signature contributions per sample (W matrix)

Proportion each signature contributes to the mutation spectrum in each individual sample

```{r child="R/contribs.Rmd"}
```
