---
title: "doomsayer_diagnostics"
output:
  html_document:
      toc: true
      toc_float: true
      keep_md: yes
      theme: united
  pdf_document:
date: '`r format(Sys.Date())`'
params:
  dev: no
  yaml_cfg: x
---

<style>
body {
    position: absolute;
    left: 0px;}
</style>

```{r setup, message=FALSE, warning=FALSE, echo=FALSE}
# knitr::opts_chunk$set(echo = TRUE)

# check and load packages
packages <- c("dplyr","tidyr","ggplot2","yaml", "devtools", "knitr", "corrr", "viridis", "heatmaply","dendextend", "e1071")

package.check <- lapply(packages, FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
        install.packages(x,
          dependencies = TRUE,
          repos = "http://cran.us.r-project.org")
        library(x, character.only = TRUE, silent=TRUE)
    }
})


if(params$dev){
  setwd("../../../Github/doomsayer/sample_output")
  yaml_cfg <- "./config.yaml"
} else {
  yaml_cfg <- params$yaml_cfg
}

# debug--paths generated from doomsayer.py refer are formatted for bash on Windows
# the sys.info check updates these paths with Windows format for proper loading in R
if(Sys.info()['sysname']=="Windows"){
  yaml_cfg <- gsub("/mnt/c", "C:", yaml_cfg)
}

yaml_args <- yaml.load_file(yaml_cfg)
if(Sys.info()['sysname']=="Windows"){
  yaml_args <- lapply(yaml_args, function(x) gsub("/mnt/c", "C:", x))
}

attach(yaml_args)
```


```{r load, echo=FALSE}
# read tables
keep_ids <- read.table(keep_path, header=F, stringsAsFactors=F)
drop_ids <- read.table(drop_path, header=F, stringsAsFactors=F)
spectra <- read.table(M_path, header=T, stringsAsFactors=F)
spectra_rates <- read.table(M_path_rates, header=T, stringsAsFactors=F)
sig_contribs <- read.table(W_path, header=T, stringsAsFactors=F)
sig_loads <- read.table(H_path, header=T, stringsAsFactors=F)
rmse <- read.table(RMSE_path, header=F, stringsAsFactors=F)
```

### Error profile heatmap

Left panel shows root mean squared error (RMSE) of each outlier sample's mutation spectrum. Right panel shows the fold-difference in contribution per subtype for each outlier. Fold-differences are calculated relative to the mean contribution of that subtype across all samples.

<!-- # ```{r heatmap, message=FALSE, warning=FALSE, echo=FALSE} -->
```{r heatmap, message=FALSE, warning=FALSE, echo=FALSE, fig.height = 18, fig.width = 14}
cutoff <- ifelse(ncol(spectra_rates)-1==6, 0.05, 0.005)
nind <- 100
names(rmse) <- c("ID", "RMSE")

sig_contribs2 <- sig_contribs[complete.cases(sig_contribs),] %>%
  # mutate(sumVar = rowSums(.[2:ncol(sig_contribs)])) %>%
  mutate(sumVar = rowSums(.[2:ncol(sig_contribs)])) %>%
  mutate_at(vars(starts_with("S")), funs(./sumVar)) %>%
  mutate(maxgp = apply(.[,2:ncol(sig_contribs)], 1, function(x)
    names(x)[which.max(x)])) %>%
  dplyr::select(-sumVar) %>%
  gather(signature, contribution, -c(ID,maxgp))

# rmse_m2 <- merge(rmse, sig_contribs2, by="ID") %>%
#   spread(signature, contribution) %>%
#   arrange(RMSE) %>%
#   mutate(pctile = ntile(RMSE, 100))

# train <- rmse_m2 %>%
#   filter(pctile<5 | pctile > 95) %>%
#   mutate(QUAL=ifelse(pctile<10, 1, 0))

# svmmod <- svm(pctile~S1+S2+S3, rmse_m2)
# rmse_m2$QUAL <- predict(svmmod, rmse_m2)

spectra_r <- spectra_rates %>%
  mutate_at(vars(-1), funs(./mean(.))) %>%
  gather(subtype, err, 2:ncol(spectra_rates)) %>%
  separate(subtype, c("category", "motif"), sep = "[.]") %>%
  mutate(err=ifelse(err>5, 5, err))

outlier_IDs <- drop_ids$V1

rmse_ord <- rmse %>%
  arrange(desc(RMSE)) %>%
  mutate(ID=factor(ID, levels=rev(ID))) %>%
  # head(nind)
  filter(ID %in% outlier_IDs)

rmse_m <- merge(rmse_ord, sig_contribs2, by="ID")
d2a <- rmse_m %>%
  gather(k1,k2,RMSE,contribution) %>%
  mutate(key=ifelse(k1=="RMSE", "RMSE", "signature \n contribution")) %>%
  mutate(NMF_out=ifelse(ID %in% drop_ids$V1, "*", ""))

top_r50 <- rmse_m %>%
  # dplyr::filter(k1 == "RMSE") %>%
  group_by(ID) %>%
  slice(which.max(contribution)) %>%
  mutate(key="RMSE") %>%
  mutate(maxgp=factor(maxgp, levels=names(sig_contribs)[-1])) %>%
  # dplyr::filter(RMSE>cutoff) %>%
  arrange(desc(RMSE)) %>%
  ungroup() %>%
  mutate(ID=factor(ID, levels=ID)) #%>%
  # head(nind)

outlier_r_m <- spectra_r %>%
  dplyr::filter(ID %in% top_r50$ID) %>%
  mutate(ID=factor(ID, levels=levels(top_r50$ID))) %>%
  mutate(subtype=paste0(category, "_", motif)) %>%
  dplyr::select(ID, subtype, err) %>%
  spread(subtype, err)

rownames(outlier_r_m) <- outlier_r_m$ID

outlier_r_m <- outlier_r_m %>%
  dplyr::select(-ID)

hrm <- heatmapr(outlier_r_m,
    k_row=3,
    Colv=TRUE,
    label_names = c("ID", "subtype", "error"),
    hclust_method="ward.D2")

# get_leaves_branches_col(hrm$rows)

hm <- heatmaply(outlier_r_m,
    k_row=3,
    Colv=TRUE,
    return_ppxpy=TRUE,
    hclust_method="ward.D2",
    colors=magma(20),
    label_names = c("ID", "subtype", "error"),
    column_text_angle = 90,
    subplot_widths=c(0.95,0.05),
    plot_method="ggplot")


rmse_ord <- rmse %>%
  arrange(desc(RMSE)) %>%
  mutate(ID=factor(ID, levels=rev(ID))) %>%
  # head(nind) %>%
  filter(ID %in% outlier_IDs) %>%
  mutate(ID=as.character(ID)) %>%
  mutate(ID=factor(ID, levels=rownames(hrm$matrix$data)))

rmse_m <- merge(rmse_ord, sig_contribs2, by="ID")
d2a <- rmse_m %>%
  gather(k1,k2,RMSE,contribution) %>%
  mutate(key=ifelse(k1=="RMSE", "RMSE", "signature \n contribution")) %>%
  mutate(NMF_out=ifelse(ID %in% drop_ids$V1, "*", ""))

rmse_plot <- rmse_m %>%
  # dplyr::filter(k1 == "RMSE") %>%
  group_by(ID) %>%
  slice(which.max(contribution)) %>%
  mutate(key="RMSE") %>%
  mutate(maxgp=factor(maxgp, levels=names(sig_contribs)[-1])) %>%
  ungroup() %>%
  mutate(ID=factor(ID, levels=rownames(hrm$matrix$data)))

plotmin <- floor(min(rmse_plot$RMSE)/0.005)*0.005

contribution_plot <- d2a %>%
  dplyr::filter(k1 != "RMSE")

breaks <- c(plotmin, plotmin/2, seq(0,abs(plotmin),abs(plotmin)/4))
labels <- c(abs(plotmin), abs(plotmin)/2, seq(0,1,0.25))

p<-ggplot(rmse_plot)+
  geom_segment(
    aes(y=ID, yend=ID,
      x=0, xend=RMSE, colour=maxgp))+
  geom_point(aes(y=ID, x=RMSE, colour=maxgp),
    size = 2) +
  scale_fill_viridis(discrete=TRUE, drop=FALSE) +
  scale_color_viridis(discrete=TRUE, drop=FALSE) +
  ylab("ID")+
  xlab("RMSE")+
  theme_bw()+
  theme(axis.text.x=element_text(size=14, angle=45, hjust=1, vjust=1),
    legend.position="none")

subplot(p, hm$p, hm$px,
    margin=0, titleX=TRUE, shareY=TRUE, widths=c(0.1, 0.85, 0.05)) %>%
  layout(autosize=F,
    width = 1400,
    height = 1400,
    margin = list(l = 130, b = 100))

top_r20 <- rmse_plot %>%
  # dplyr::filter(RMSE>cutoff) %>%
  arrange(desc(RMSE)) %>%
  ungroup() %>%
  mutate(ID=factor(ID, levels=ID)) %>%
  head(20)
```

### Combined mutation spectrum

These plots shows the total number of observations in each subtype, combined across all samples

#### All samples

```{r dist, echo=FALSE, fig.height = 12, fig.width = 24}
# get_leaves_branches_col(hrm$rows)

# overall distribution
spectra2 <- spectra %>%
  mutate(group=ifelse(ID %in% keep_ids$V1, "keep", "drop")) %>%
  group_by(group) %>%
  gather(subtype, count, 2:ncol(spectra)) %>%
  separate(subtype, c("category", "motif"), sep = "[.]") %>%
  group_by(category, motif, group) %>%
  summarise(count=sum(count)) %>%
  group_by(group) %>%
  mutate(prop=count/sum(count)) %>%
  ungroup()

ggplot(spectra2, aes(x=motif, y=prop, fill=group, group=group))+
  geom_bar(stat="identity", position="dodge")+
  facet_wrap(~category, ncol=6, scales="free_x")+
  ylab("Contribution")+
  scale_fill_brewer(palette = "Set1")+
  theme_bw()+
  theme(axis.text.x=element_text(angle=90, vjust=0, size=20),
        axis.text.y=element_text(size=20),
        axis.title.x=element_text(size=24),
        axis.title.y=element_text(size=24),
        legend.title=element_blank(),
        legend.text=element_text(size=20),
        strip.text=element_text(size=24))
        # legend.position="none")
```

### Signature loadings (H matrix)

Describes how each mutation subtype is loaded into the r signatures

```{r sigloads, message=FALSE, warning=FALSE, echo=FALSE, fig.height = 6, fig.width = 14}
# signature loadings
corrm <- sig_loads %>%
  gather(var, val, 2:ncol(sig_loads)) %>%
  spread(Sig, val) %>%
  dplyr::select(-var) %>%
  correlate

corrm

sig_loads_long <- sig_loads %>%
  mutate(sumVar = rowSums(.[2:ncol(sig_loads)])) %>%
  rowwise() %>%
  mutate_each(funs(./sumVar), -Sig) %>%
  gather(subtype, loading, 2:ncol(sig_loads)) %>%
  separate(subtype, c("category", "motif"), sep = "[.]")

ggplot(sig_loads_long, aes(x=motif, y=loading, fill=Sig))+
  geom_bar(stat="identity")+
  facet_grid(Sig~category, scales="free_x")+
  scale_fill_viridis(discrete=TRUE)+
  theme_bw()+
  theme(axis.text.x=element_text(angle=90, vjust=0, size=10),
        axis.text.y=element_text(size=20),
        axis.title.x=element_text(size=24),
        axis.title.y=element_text(size=24),
        strip.text=element_text(size=24),
        legend.position="none")
```

### Signature contributions per sample (W matrix)

Proportion each signature contributes to the mutation spectrum in each individual sample

```{r sigs, message=FALSE, warning=FALSE, echo=FALSE, fig.height = 12, fig.width = 10}

contribution_plot <- sig_contribs2 %>%
  filter(ID %in% rmse_plot$ID) %>%
  spread(signature, contribution) %>%
    mutate(ID=factor(ID, levels=levels(rmse_plot$ID)))

pc <- plot_ly(contribution_plot, x=~S1, y=~ID, type = 'bar', orientation = 'h', marker=list(color=viridis(3)[1])) %>%
  add_trace(x = ~S2, marker=list(color=viridis(3)[2])) %>%
    add_trace(x = ~S3, marker=list(color=viridis(3)[3])) %>%
  layout(barmode = 'stack',
    showlegend = FALSE)

pc
```
