---
title: "doomsayer_mutation_signature_analysis"
output:
  html_document:
      toc: true
      toc_float: true
      keep_md: yes
      theme: united
  pdf_document:
date: '`r format(Sys.Date())`'
params:
  dev: no
  yaml_cfg: x
---

<style>
body {
    position: absolute;
    left: 0px;}
</style>

```{r setup, message=FALSE, warning=FALSE, echo=FALSE}
# knitr::opts_chunk$set(echo = TRUE)

# check and load packages
packages <- c("dplyr","tidyr","ggplot2","yaml", "devtools", "knitr", "corrr", "viridis", "heatmaply","dendextend", "e1071")

package.check <- lapply(packages, FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
        install.packages(x,
          dependencies = TRUE,
          repos = "http://cran.us.r-project.org")
        library(x, character.only = TRUE, silent=TRUE)
    }
})

if(params$dev){
  setwd("../../../Github/doomsayer/sample_output")
  yaml_cfg <- "./config.yaml"
} else {
  yaml_cfg <- params$yaml_cfg
}

# debug--paths generated from doomsayer.py refer are formatted for bash on Windows
# the sys.info check updates these paths with Windows format for proper loading in R
if(Sys.info()['sysname']=="Windows"){
  yaml_cfg <- gsub("/mnt/c", "C:", yaml_cfg)
}

yaml_args <- yaml.load_file(yaml_cfg)
if(Sys.info()['sysname']=="Windows"){
  yaml_args <- lapply(yaml_args, function(x) gsub("/mnt/c", "C:", x))
}

attach(yaml_args)
```

```{r load, echo=FALSE}
# read tables
# keep_ids <- read.table(keep_path, header=F, stringsAsFactors=F)
# drop_ids <- read.table(drop_path, header=F, stringsAsFactors=F)
spectra <- read.table(M_path, header=T, stringsAsFactors=F)
spectra_rates <- read.table(M_path_rates, header=T, stringsAsFactors=F)
sig_contribs <- read.table(W_path, header=T, stringsAsFactors=F)
sig_loads <- read.table(H_path, header=T, stringsAsFactors=F)
rmse <- read.table(RMSE_path, header=F, stringsAsFactors=F)
```

### Signature contributions per sample (W matrix)

Proportion each signature contributes to the mutation spectrum in each individual sample

<!-- # ```{r heatmap, message=FALSE, warning=FALSE, echo=FALSE} -->
```{r sigs, message=FALSE, warning=FALSE, echo=FALSE, fig.height = 18, fig.width = 14}
cutoff <- ifelse(ncol(spectra_rates)-1==6, 0.05, 0.005)
nind <- 100
names(rmse) <- c("ID", "RMSE")

sig_contribs2 <- sig_contribs[complete.cases(sig_contribs),] %>%
  # mutate(sumVar = rowSums(.[2:ncol(sig_contribs)])) %>%
  mutate(sumVar = rowSums(.[2:ncol(sig_contribs)])) %>%
  mutate_at(vars(starts_with("S")), funs(./sumVar)) %>%
  mutate(maxgp = apply(.[,2:ncol(sig_contribs)], 1, function(x)
    names(x)[which.max(x)])) %>%
  dplyr::select(-sumVar) %>%
  gather(signature, contribution, -c(ID,maxgp))

ggplot(sig_contribs2, aes(x=ID, y=contribution, fill=signature))+
  geom_bar(stat="identity", position="stack")+
  facet_wrap(~maxgp, scales="free")+
  theme_bw()+
  theme(axis.text.x=element_blank())

```

```{r sigs2, message=FALSE, warning=FALSE, echo=FALSE, fig.height = 12, fig.width = 10}

# contribution_plot <- sig_contribs2 %>%
#   filter(ID %in% rmse_plot$ID) %>%
#   spread(signature, contribution) %>%
#     mutate(ID=factor(ID, levels=levels(rmse_plot$ID)))
#
# pc <- plot_ly(contribution_plot, x=~S1, y=~ID, type = 'bar', orientation = 'h', marker=list(color=viridis(3)[1])) %>%
#   add_trace(x = ~S2, marker=list(color=viridis(3)[2])) %>%
#     add_trace(x = ~S3, marker=list(color=viridis(3)[3])) %>%
#   layout(barmode = 'stack',
#     showlegend = FALSE)
#
# pc
```

### Combined mutation spectrum

These plots shows the total number of observations in each subtype, combined across all samples

#### All samples

```{r dist, echo=FALSE, fig.height = 12, fig.width = 24}
# get_leaves_branches_col(hrm$rows)

# overall distribution
spectra2 <- spectra %>%
  mutate(group="all") %>%
  group_by(group) %>%
  gather(subtype, count, 2:ncol(spectra)) %>%
  separate(subtype, c("category", "motif"), sep = "[.]") %>%
  group_by(category, motif, group) %>%
  summarise(count=sum(count)) %>%
  group_by(group) %>%
  mutate(prop=count/sum(count)) %>%
  ungroup()

ggplot(spectra2, aes(x=motif, y=prop, fill=group, group=group))+
  geom_bar(stat="identity", position="dodge")+
  facet_wrap(~category, ncol=6, scales="free_x")+
  ylab("Contribution")+
  scale_fill_brewer(palette = "Set1")+
  theme_bw()+
  theme(axis.text.x=element_text(angle=90, vjust=0, size=20),
        axis.text.y=element_text(size=20),
        axis.title.x=element_text(size=24),
        axis.title.y=element_text(size=24),
        legend.title=element_blank(),
        legend.text=element_text(size=20),
        strip.text=element_text(size=24))
        # legend.position="none")
```

### Signature loadings (H matrix)

Describes how each mutation subtype is loaded into the r signatures

```{r sigloads, message=FALSE, warning=FALSE, echo=FALSE, fig.height = 6, fig.width = 14}
# signature loadings
corrm <- sig_loads %>%
  gather(var, val, 2:ncol(sig_loads)) %>%
  spread(Sig, val) %>%
  dplyr::select(-var) %>%
  correlate

corrm

sig_loads_long <- sig_loads %>%
  mutate(sumVar = rowSums(.[2:ncol(sig_loads)])) %>%
  rowwise() %>%
  mutate_each(funs(./sumVar), -Sig) %>%
  gather(subtype, loading, 2:ncol(sig_loads)) %>%
  separate(subtype, c("category", "motif"), sep = "[.]")

ggplot(sig_loads_long, aes(x=motif, y=loading, fill=Sig))+
  geom_bar(stat="identity")+
  facet_grid(Sig~category, scales="free_x")+
  scale_fill_viridis(discrete=TRUE)+
  theme_bw()+
  theme(axis.text.x=element_text(angle=90, vjust=0, size=10),
        axis.text.y=element_text(size=20),
        axis.title.x=element_text(size=24),
        axis.title.y=element_text(size=24),
        strip.text=element_text(size=24),
        legend.position="none")
```
